---
title: "Week 4. Spatial Demography"
author: "Myunggu Jung"
format: html
editor: visual
---

## Mapping with \`tmap\`

In this practice, we learn how to create thematic maps using the \`tmap\` package in R. \
The package is introduced from the official tmap website: \<https://r-tmap.github.io/tmap/\>.\
\
More examples, tutorials, and practice exercises are available on the website.

### 1. Install and Load packages

If you are running this code for the first time, install the required packages.

```{r}
#| eval: false

install.packages("tmap")
  
```

We now open the tmap library.

```{r}
#| warning: false
#| message: false
library(tmap)
```

### 2. Set tmap mode

The \`tmap\` package has two main modes.\
1) \`plot\` mode creates static maps for slides, reports, and academic papers.

2\) \`view\` mode creates interactive maps using leaflet.

```{r}
#| warning: false
#| message: false

tmap_mode("plot")
tmap_mode("view")

```

### 3. Load example spatial data

World is an example sf object included in the tmap package.

The dataset includes country boundaries and several demographic, health, and socioeconomic indicators.

```{r}
#| warning: false
#| message: false

data(World, package = "tmap")

# Check available variables
names(World)
```

### 4. Basic map

A basic map colours geographic areas according to the value of a variable.

```{r}
#| warning: false
#| message: false
tm_shape(World) +
  tm_polygons(fill = "HPI")

# Here, countries are coloured by the Happy Planet Index.
```

### 5. Basic map with colour palette

We can map life expectancy using a continuous colour scale.

```{r}
#| warning: false
#| message: false

tm_shape(World) +
  tm_polygons(
    fill = "life_exp",
    fill.scale = tm_scale_continuous(values = "viridis"),
    fill.legend = tm_legend(
      title = "Life Expectancy",
      orientation = "landscape",
      frame = FALSE
    )
  )

# Common palettes:
# "YlGn", "YlGnBu", "GnBu", "BuGn", "PuBu", "PuBuGn", "BuPu"
# "Blues", "Greens", "Reds", "Oranges", "Purples"
# "RdYlBu", "RdYlGn", "RdBu", "Spectral"
# "viridis", "plasma", "magma", "inferno", "cividis"
```

### 6. Map with fixed class intervals

Here, life expectancy is divided into fixed groups: 0–60, 60–70, 70–80, and 80–90 years

```{r}
#| warning: false
#| message: false

le <- tm_shape(World) +
  tm_polygons(
    fill = "life_exp",
    fill.scale = tm_scale_intervals(
      style = "fixed",
      breaks = c(0, 60, 70, 80, 90),
      values = "brewer.yl_gn_bu"
    ),
    fill.legend = tm_legend(
      title = "Life Expectancy",
      orientation = "landscape",
      frame = FALSE
    )
  )

le
```

Fixed intervals make the map easier to interpret when the categories have substantive meaning.

### 7. Add transparency

Transparency can be controlled using \`fill_alpha\`.

\`fill_alpha = 1.0\` is fully opaque, while \`fill_alpha = 0.0\` is fully transparent.

```{r}
#| warning: false
#| message: false
le_a <- tm_shape(World) +
  tm_polygons(
    fill = "life_exp",
    fill_alpha = 0.5,
    fill.scale = tm_scale_intervals(
      breaks = c(0, 60, 70, 80, 90),
      values = "brewer.yl_gn_bu"
    ),
    fill.legend = tm_legend(
      title = "Life Expectancy"
    )
  )

le_a
```

### 8. Interactive map

We can switch to interactive mode. Interactive maps are useful for exploratory spatial analysis.\
We return to static plot mode for the rest of the lecture.

```{r}
#| warning: false
#| message: false

tmap_mode("view")

le_a

tmap_mode("plot")

le_a

```

### 8. Multiple maps

Multiple maps allow us to compare indicators side by side.

```{r}
#| warning: false
#| message: false

tm_shape(World) +
  tm_polygons(
    fill = c("well_being", "life_exp"),
    fill.legend = tm_legend("")
  ) +
  tm_layout(
    panel.labels = c("A. Well-being", "B. Life Expectancy")
  )

```

This helps us compare whether countries with high life expectancy also show high well-being.

```{r}
#| warning: false
#| message: false

tm_shape(World) +
  tm_polygons(
    fill = c("well_being", "life_exp", "pop_est", "economy")
  )

```

This helps us compare whether countries with high life expectancy also show high well-being ,population estimation, Economy.

### 9. Faceted Maps

In many demographic studies, we are interested not only in the overall spatial pattern but also in how patterns differ across regions. One useful approach is **faceting**, which creates multiple maps using the same variable while dividing the data into meaningful groups.

In this example, we use municipality-level data from the Netherlands and create separate maps for each province.

The `NLD_muni` dataset contains municipality boundaries and several socioeconomic indicators.

```{r}
#| warning: false
#| message: false

data(NLD_muni, package = "tmap") 

names(NLD_muni)
  

```

First, we create a single choropleth map showing employment rates across all municipalities.

The `NLD_muni` dataset contains municipality boundaries and several socioeconomic indicators.

```{r}
#| warning: false
#| message: false

tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate")

```

This map shows the national spatial distribution of employment rates.

We now split the map by province using `tm_facets()`.

```{r}
#| warning: false
#| message: false

tm_shape(NLD_muni) +
  tm_polygons(fill = "employment_rate") +
  tm_facets(by = "province")

```

The argument `by = "province"` tells tmap to create a separate map for each province.

Instead of displaying one large national map, the output contains multiple smaller maps arranged in panels.

## Practice Exercise

### Exercise 1. Comparing Population Age Structure

Create a set of maps showing the spatial distribution of the following age groups across Dutch municipalities:

-   Population aged 0–14 years

-   Population aged 15–24 years

-   Population aged 25–44 years

-   Population aged 65 years and over

### Exercise 2. Faceted Maps by Province

Each student will select one indicator and create a faceted map by province.

Available indicators:

-   Student 1: Employment Rate

-   Student 2: Population Aged 0–14 Years

-   Student 3: Population Aged 25–44 Years

-   Student 4: Population Aged 65 Years and Over

-   Student 5: Urbanity
