---
title: "Week 2. Comtemporary Mortality and Fertility Transition"
author: "Myunggu Jung"
format: html
editor: visual
---

## Part 1: Lecture Slide

You can download the lecture slide here: [Week2_Lecture_Slide](https://github.com/myunggu82/global-health-demography/blob/main/lectures/week_02_0609.pdf)\
And recap exercise: [Week2_Exercise](https://github.com/myunggu82/global-health-demography/blob/main/lectures/exercise.pptx)

## Part 2: Practice: Mortality and Fertility

### 1. Objectives

By the end of this lecture, students will be able to:\
\
1. Caculate and compare Crude death rate and Age-standardized mortality rates.\
2. Understand changes in survival curve.\
3. Calculate and compare changes in Age-specific fertility rates

### 2. Install and Load packages

If you are running this code for the first time, install the required packages.

```{r}
#| eval: false

install.packages("DemoTools")
  

```

We now load the required packages for data wrangling, visualization, and animation.

```{r}
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(DemoTools)
library(wpp2024)

```

### 3. Load WPP 2024 population and mortality data

The \`wpp2024\` package contains population, fertility, mortality, and projection datasets from the UN World Population Prospects 2024.

```{r}
#| warning: false
#| message: false
data(mx1dt)
data(mx5dt)
data(popAge1dt)
```

### 4.  Age-specific mortality and population (Japan vs Uganda, 2018)

This section compares mortality between Japan and Uganda using age-specific mortality rates and population data.

```{r}
#| warning: false
#| message: false
jap_mx <- mx1dt %>%
  filter(name == "Japan", year == 2018) %>%
  select(age, mxB)

uga_mx <- mx1dt %>%
  filter(name == "Uganda", year == 2018) %>%
  select(age, mxB)

jap_pop <- popAge1dt %>%
  filter(name == "Japan", year == 2018) %>%
  select(age, pop)

uga_pop <- popAge1dt %>%
  filter(name == "Uganda", year == 2018) %>%
  select(age, pop)
```

we first calculate 'Crude Death Rate'.

```{r}
#| warning: false
#| message: false

jap_cdr <- jap_pop %>%
  left_join(jap_mx, by = "age") %>%
  summarise(CDR = sum(pop * mxB) / sum(pop) * 1000) %>%
  pull(CDR)

uga_cdr <- uga_pop %>%
  left_join(uga_mx, by = "age") %>%
  summarise(CDR = sum(pop * mxB) / sum(pop) * 1000) %>%
  pull(CDR)

```

The crude death rate shows the number of deaths per 1,000 population without adjusting for age structure.

### 5.  Direct age standardization

```{r}
#| warning: false
#| message: false
std_pop <- jap_pop %>%
  rename(pop_jap = pop) %>%
  left_join(uga_pop %>% rename(pop_uga = pop), by = "age") %>%
  mutate(std_pop = (pop_jap + pop_uga) / 2) %>%
  select(age, std_pop)

jap_asmr <- std_pop %>%
  left_join(jap_mx, by = "age") %>%
  summarise(ASMR = sum(std_pop * mxB) / sum(std_pop) * 1000) %>%
  pull(ASMR)

uga_asmr <- std_pop %>%
  left_join(uga_mx, by = "age") %>%
  summarise(ASMR = sum(std_pop * mxB) / sum(std_pop) * 1000) %>%
  pull(ASMR)

```

Age-standardized mortality removes the effect of different population age structures.

The below table shows how crude and age-standardized mortality rates can lead to different interpretations.

```{r}
#| warning: false
#| message: false
mortality_result <- tibble(
  Country = c("Japan", "Uganda"),
  `Crude Death Rate` = c(jap_cdr, uga_cdr),
  `Age-standardized Mortality Rate` = c(jap_asmr, uga_asmr)
)

mortality_result

```

### 6.  Survivorship curves: Republic of Korea, 1953 vs 2018

We first need to create time tables and the use lx to create survivoership curves. Survivorship curves show how many people remain alive at each age out of 100,000 births.

```{r}
#| warning: false
#| message: false

kor_mx_1953 <- mx5dt %>%
  filter(name == "Republic of Korea", year == 1953) %>%
  transmute(
    Age = as.numeric(gsub("100\\+", "100", age)),
    nMx = mxB
  ) %>%
  arrange(Age)

kor_mx_2018 <- mx5dt %>%
  filter(name == "Republic of Korea", year == 2018) %>%
  transmute(
    Age = as.numeric(gsub("100\\+", "100", age)),
    nMx = mxB
  ) %>%
  arrange(Age)

lt_kor_1953 <- lt_abridged(
  nMx = kor_mx_1953$nMx,
  Age = kor_mx_1953$Age,
  radix = 100000
) %>%
  mutate(year = "1953")

lt_kor_2018 <- lt_abridged(
  nMx = kor_mx_2018$nMx,
  Age = kor_mx_2018$Age,
  radix = 100000
) %>%
  mutate(year = "2018")

lt_kor_both <- bind_rows(lt_kor_1953, lt_kor_2018)
```

We now plot survivorship curves

```{r}
ggplot(
  lt_kor_both,
  aes(x = Age, y = lx, color = year, linetype = year)
) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(
    values = c("1953" = "red", "2018" = "blue")
  ) +
  labs(
    title = "Survivorship Curves, Republic of Korea",
    subtitle = "Comparison between 1953 and 2018",
    x = "Age",
    y = "Survivors out of 100,000 births",
    color = "Year",
    linetype = "Year"
  ) +
  theme_classic()
```

### 6.  Fertility Transition in Korea

This section examines changes in age-specific fertility rates in the Republic of Korea.
The results show how fertility shifted from a broad and high-fertility pattern to a delayed and low-fertility pattern.

```{r}
#| warning: false
#| message: false
data(percentASFR5dt) 
data(tfr5dt)

kor_asfr <- percentASFR5dt %>%
  filter(
    name == "Republic of Korea",
    year %in% c(1953, 1973, 1993, 2013, 2018)
  ) %>%
  left_join(
    tfr5dt %>%
      select(name, year, tfr),
    by = c("name", "year")
  ) %>%
  mutate(
    ASFR = tfr * pasfr / 500
  )

kor_asfr
```

The percentage distribution of fertility by age is converted into age-specific fertility rates using the total fertility rate.

```{r}
kor_asfr %>%
  group_by(year) %>%
  summarise(
    TFR_check = 5 * sum(ASFR),
    TFR_original = first(tfr)
  )

```

This check confirms that the calculated age-specific fertility rates reproduce the original total fertility rate.

The percentage distribution of fertility by age is converted into age-specific fertility rates using the total fertility rate.

```{r}
kor_asfr_table <- kor_asfr %>%
  mutate(
    ASFR = round(ASFR * 1000, 1)
  ) %>%
  select(year, age, ASFR) %>%
  pivot_wider(
    names_from = age,
    values_from = ASFR
  ) %>%
  arrange(year)

kor_asfr_table

```

The table shows age-specific fertility rates per 1,000 women.

fertility rate.

```{r}
ggplot(
  kor_asfr,
  aes(
    x = age,
    y = ASFR * 1000,
    color = factor(year),
    group = year
  )
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.5) +
  scale_color_viridis_d(
    option = "viridis",
    begin = 0.1,
    end = 0.9
  ) +
  labs(
    title = "Age-Specific Fertility Rates, Republic of Korea",
    subtitle = "1953, 1973, 1993, 2013, and 2018",
    x = "Age",
    y = "Births per 1,000 women",
    color = "Year"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom"
  )

```

The fertility schedule became much lower and shifted toward older ages over time.\
This reflects Korea's rapid fertility transition and delayed childbearing.
